A Formalization of Complete Discrete Valuation Rings and Local Fields

### 2023-09-19 CPP_paper/CPP_paper.tex
added CPP_paper folder and graph

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[A Formalization of Complete DVRs and Local Fields]{A Formalization of Complete Discrete Valuation Rings and Local Fields -- OLD ONE, FORGET THIS FILE}

@@

\section{Introduction}
\begin{itemize}
\item Say all rings are commutative and unitary.
\item Say something, or give a reference, to the role played by ``type classes'', ``instances'' and class-type inference in \lean and/or \mathlib.
\item Remember to create the graph
\item Acknowledge de Frutos-Fernández and Yaël Dillies for allowing us to use their work.
\item Update link to repository \href{https://github.com/mariainesdff/local_class_field_theory}{\extlink}.
\item Mention dot notation when discussing Lean and mathlib.

% \mi{Links are rendered in black with the "review" option, which means they will probably be missed. In the sphere eversion paper they have a symbol next to each link; do you know how to achieve this? Otherwise we could ask them.}\fil{I added this and you can use the \code{extlink} command (see what I did for the \code{valuation} structure). Yet, in that case (see my comment there) I would also add the corresponding lean code, no? Now that it is available, do we want to systematically add this external link to \mathlib, though, even when we are simply mentioning the library, as in \mathlib\href{https://leanprover-community.github.io/mathlib_docs/}{\extlink}?}
\end{itemize}
\section {Discrete Valuation Rings}\label{section:dvr}

\subsection{Valuations}\label{subsection:valuations}
Let $R$ be a ring \mi{and let $\Gamma_0$ be a linearly ordered commutative monoid with zero, that is, a monoid with a partial order in which any two elements are comparable, and an added zero element that is smaller than all the others}. A \emph{valuation} $v$ on $R$ is a map $v \colon R \to \Gamma_0$ such that
\begin{enumerate}
	\item $v (0) = 0$,
	\item $v (1) = 1$,
	\item $v (x \cdot y) = v (x) \cdot v (y)$ for all $x, y \in R$, and
	\item $v (x + y) \le \max {v (x), v (y)}$ for all $x, y \in R$.
\end{enumerate}

For instance, when $R$ is a Dedekind domain of Krull dimension 1 with fraction field $K$ and $\m$ is a maximal ideal of $R$, we can define a function $c_\m \colon R \to \ZZ \cup \{ \infty\}$ \fil{we need to say that ``monoiod with $0$`` or ``monoid with infinity`` are the same thing...} by sending zero to $\infty$ and any nonzero $r \in R$ to the number of times that $\m$ appears in the factorization of the principal ideal $(r)$. Note that this number is well defined because ideals in a Dedekind domain admit a unique factorization into maximal ideals, up to reordering. We can extend the function $c_\m$ to $K$ by setting $c_\m (r/s) := c_\m (r) - c_\m (s)$ for any $r, s \in R$ with $s \ne 0$.

Then, for any real number $n_\m > 1$, it is easy to check that the function
$ v_{\m}\colon K \to n_\m^{\mathbb{Z} \cup \{-\infty\}} = n_\m^\mathbb{Z} \cup \{0\}$ sending $x$ to $n_\m^{-c_\m(x)}$ is a valuation on $K$, which we call the \textit{$\m$-adic valuation}. For example, when $R = \ZZ$, $K = \QQ$ and $\m = (p)$ is the ideal generated by a prime number $p$, $v_p$ is the $p$-adic (multiplicative) valuation on $\QQ$, also known as the $p$-adic absolute value.

Note that the choice of number $n_\m > 1$ in the definition of $v_\m$ is not relevant, in the sense that any two such choices will give rise to equivalent valuations. In our formalization, we abstract the codomain $n_\m^\mathbb{Z} \cup \{0\}$ using the type
\texttt{with\_zero (multiplicative $\mathbb{Z}$)}, denoted $\wzmZZ$,
which carries the multiplicative structure corresponding to the additive structure on $\ZZ$, as well as an added zero element.

For example, the rational numbers $20$ and $25$ have $5$-adic valuation $(-1 : \wzmZZ)$ and $(-2 : \wzmZZ)$, respectively, so $25$ is smaller than $20$ in the $p$-adic topology. \mi{Do you want to add anything else about this?}\fil{Probably, in the previous paragraph, I would add some of the main properties of $\wzmZZ$: something like $2+3=6$, that there are three elements $0:\wzmZZ,1:\wzmZZ=$\code{to_add}$0:\ZZ$ and \code{to_add}$1:\ZZ$ that are different, and some formulae with inequalities. Observe that. for our definition of uniformizer, the term \code{to_add}$-1:\ZZ$ is essential.}

We represent valuations in Lean using the \code{valuation} structure \href{https://leanprover-community.github.io/mathlib_docs/ring_theory/valuation/basic.html#valuation}{\extlink}
 available in \mathlib, which encodes our above definition. A valuation on a ring $R$ induces a topology on $R$ for which the ring operations are continuous, hence turning $R$ into a topological ring.
Mathlib also provides the class \href{https://leanprover-community.github.io/mathlib_docs/topology/algebra/valuation.html#valued}{valued}\fil{why not inserting the}
\begin{lstlisting}
class (R : Type*) [ring R] valued : extends...
\end{lstlisting}
\fil{?}
, which bundles together a ring with a
uniform space structure and a distinguished valuation that induces it. This class is designed for rings in which there is a canonical valuation. For example, on the field $\QQ_p$ of $p$-adic numbers, there is a unique nontrivial valuation, which we register as an instance of the valued class. By contrast, on the rational numbers there are infinitely many valuations, none of them preferred over the others, so we do not put a global valued instance on $\QQ$.

Given any ring $R$ with a valuation $v$, we can define the \textit{unit ball} of $R$ as the subring $\ball{R}$ of elements with valuation less than or equal to one. In \mathlib, this subring is called \code{valuation.integer v} \href{https://leanprover-community.github.io/mathlib_docs/ring_theory/valuation/integers.html#valuation.integer}{\extlink}. \mi{Moreover, when $K$ is a field, then its subring $\ball{K}$ has the property that for every element $k$ of $K$, either $k \in \ball{K}$ or $k^{-1} \in \ball{K}$. Such a subring is called a \emph{valuation subring}. Note that the definition of valuation subring involves taking inverses, and hence is only available for fields. Therefore, when working with a general ring $R$, we will formalize $\ball{R}$ as \code{v.integer}, but when working with a field $K$ we will use the stronger definition \code{v.valuation_subring}, which gives us access to extra results about valuation fields available in mathlib} \href{https://leanprover-community.github.io/mathlib_docs/ring_theory/valuation/valuation_subring.html#valuation_subring}{\extlink}.

\mi{TODO} This should not be confused with \mathlib's \code{valuation\_ring} \href{https://leanprover-community.github.io/mathlib_docs/ring_theory/valuation/valuation_ring.html#valuation_ring}{\extlink} class \fil{should we explain the difference and the relation with this latter class with the \code{valuation_subring} one?}

\mi{
\begin{itemize}
%	\item Properties of valuations valued in $\wzmZZ$, together with the \code{multiplicative} and \texttt{with\_zero} classes, and clashes with "usual" intuition. Keep in mind that \emph{with\_zero multiplicative} $\ZZ$ is ``the same'' as $q^{-\ZZ}\cup \{0\}$.
%	\item Difference between \texttt{valued} (which is a class, and carries a topology/uniformity) and \texttt{valuation} that is just a function.
%	\item We put \texttt{valued} instances on fields but "never" on rings that are not fields.
%	\item There is something called \texttt{valuation\_ring} that has (almost) nothing to do; but \texttt{valuation\_subring} has.
	\item Comment about the choice of \texttt{.integers} instead of \texttt{.valuation\_subring}
\end{itemize}
}
\subsection{Definition of Discrete Valuation Rings}\label{subsection:def_dvr}
Our main reference for the general theory of discrete valuation rings are~\cite{Ser62} and~\cite[Chapitre~VI]{Bou85}. We refer to~\cite{Bou07}, in particular to \cite[Chapitre~IV]{Bou07}, for the main results about ring theory and commutative algebra that we will need.

Let $R$ be a ring, \fil{as above assumed} commutative and unitary. We begin by recalling the following equivalence
\begin{theorem}[{\cite[Chapitre~VI, \S3, n$^\circ$~6, Proposition~8]{Bou85} and~\cite[Chapitre I, \S2, Propositions~2 and~3]{Ser62}}]\label{thm:dvr_tfae}
The following properties are equivalent:
\begin{listResults}
\item $R$ is a Noetherian local ring whose maximal ideal $\maxid$ is generated by a non-nilpotent element;
\item $R$ is a Noetherian local integral domain that is integrally closed and is not a field;
\item $R$ is a local PID that is not a field;
\end{listResults}
\end{theorem}
\begin{remark}
In~\cite[Chapitre~VI, \S3, n$^\circ$ 6, Proposition~8]{Bou85} other equivalences are proved, but we will not need them.
\end{remark}
A ring satisfying the equivalent properties of Theorem~\ref{thm:dvr_tfae} is called a discrete valuation ring (often shortened as DVR). The definition is motivated by the fact that a DVR $R$ can always be endowed with a class of valuations, 
\[
v\colon R\longrightarrow \ZZ\cup\{+\infty\}\fil{\text{ or use the mathlib convention?}}.
\]
For each of these valuations, the image $\image(v)\subseteq\ZZ\cup \{+\infty\}$ is the free monoid generated by $v(r)$, where $r$ is any generator of the maximal ideal $\maxid$ of $R$. Upon replacing $v$ by $v/v(r)$, for some generator $r$ of $\maxid$, we can assume that this free monoid coincides with $\NN$: in this case the valuation is said to be normalised, and the elements of valuation equal to $1$ are called uniformizers. One basic result is that, for a normalised valuation, the uniformisers are exactly the set of generators of $\maxid$, \fil{because \dots}. Moreover, there exists a unique normalised valuation on a DVR, since every valuation is uniquely determined by its value on the generators of $\maxid$. We denote it by ${\normalised}_{,R}$ --- or simply by $\normalised$ when the DVR $R$ can be understood.

This point of view differs slightly from the classical one (as found, for instance, in~\cite[Chapitre~I]{Ser62} or in~\cite[???]{Bou85}), who \fil{only FALSE} consider the normalised valuation on a DVR, and regard it as being ``canonical''. This should not come as a surprise in the setting of a formalisation work: on the one hand, regarding ``the'' valuation as part of the data would \fil{rigidify the situation too much NOT NICE}, as in \fil{give a REF to some discussion in our DD paper about fraction fields} and, on the other, there might be different constructions of a valuation on a ring that would turn out to be propositionally equal albeit not definitionally equal. The corresponding pairs $(R,v)$, for the several possible constructions of $v$, would all be mathematically intercheangable yet represented by different, although equivalent types; this would lead to unnecessary verbosity while keeping track of the equivalences: in \fil{\S\dots and in \S\dots we will see explicit examples of this phenomenon, supporting our decision of avoiding the choice of a preferred valuation on $R$}. In particular, the definition we propose of a discrete valuation is encoded in the following class:
\begin{lstlisting}
class is_discrete (v : valuation A $\wzmZZ$) : Prop := (surj : function.surjective v)
\end{lstlisting}
This \fil{differs} from the usual definition (see, for instance~\cite[Proposition~I.1]{Ser62}), where one declares a valuation to be discrete when its image is isomorphic to a subgroup of $\ZZ$ (enriched with an element $\infty$), and observes that such a valuation can always be normalized to become surjective. \fil{say that we rigidify things by forcing the map $v$ to be $\wzmZZ$-valued}. In particular, with our definition, a valuation is discrete if and only if there exists a uniformizer, which we prove in the following form:
\begin{lstlisting}
lemma is_discrete_of_exists_uniformizer {K : Type*} [field K] (v : valuation K $\wzmZZ$) {$\pi$ : K} (h$\pi$ : is_uniformizer v $\pi$) : is_discrete v := sorry
\end{lstlisting}
The lemma \code{exists_uniformizer_of_discrete} provides the reverse implication . Similarly, the aforementioned correspondance between uniformizers for a normalised valuation and generators of $\maxid$ takes the form
\begin{lstlisting}
lemma uniformizer_is_generator ($\pi$ : uniformizer v) : maximal_ideal v.valuation_subring = ideal.span {$\pi.1$} := sorry
\end{lstlisting}
where the fact that the valuation is normalised becomes \emph{a consequence}, rather than an assumption as in the classical approach. For the same reason, in the declaration formalising the opposite implication, namely
\begin{lstlisting}
lemma is_uniformizer_of_generator {r : K$_0$} [v.is_discrete]
  (hr : maximal_ideal v.valuation_subring = ideal.span {r}) : is_uniformizer v r :=
\end{lstlisting}
the assumption that \code{v} is discrete --- implemented as the \code{class} assumption \code{[v.is_discrete]} --- becomes necessary.

Given a DVR, that we implement in \lean using the \mathlib class \code{discrete_valuation_ring}\href{https://leanprover-community.github.io/mathlib_docs/ring_theory/discrete_valuation_ring/basic.html#discrete_valuation_ring}{extlink} --- this \fil{corresponds to property...??? in the above theorem} --- we can consider the adic valuation $\normalised$ attached to its maximal ideal, \fil{as discussed above}. We produce an \fil{check A vs R}
\begin{lstlisting}
instance discrete_valuation.is_discrete : (@valued.v (fraction_ring A) _ $\wzmZZ$ _ _) := sorry
\end{lstlisting}
showing that $\normalised$ is discrete according to our definition. Anticipating subsequent applications to algebraic number \fil{theory and to a global-to-local approach that\dots}, we also prove the 
\begin{lstlisting}
lemma (R : Type*) [comm_ring R] [is_domain R] [is_dedekind_domain R] (K : Type*) [field K] [algebra R K] [is_fraction_ring R K] ($\maxid$ : height_one_spectrum R) is_dedekind_domain.height_one_spectrum.adic_valuation.is_discrete : is_discrete (adic_valued $\maxid$).v := 
\end{lstlisting}
It shows that starting with a Dedekind domain $R$ \fil{GIVE REF}, a maximal ideal $\maxid$ in it, the associated $\maxid$-adic valuation is discrete with respect to our definition and therefore defines a \code{valued} structure on the field $K=\Frac(R)$. {As an application, in this setting, both the localization $R_{\maxid}$ and the completion $\widehat{R_{\maxid}}$ get a structure of a DVR}.

Since a DVR is an integral domain, it is possible to extend the valuation to its field of fractions $K=\Frac(R)$ by setting $v(a/b)=v(b)-v(a)$ \fil{or...}. If the valuation $v$ on $R$ is normalised, we obtain a \emph{surjective} valuation
\[
K\longtwoheadrightarrow \ZZ\cup\{+\infty\}\text{\fil{or\dots}}
\]
whose kernel coincides with $\unit{R}$. \fil{With the definitions introduced in} \S\ref{subsection:valuations}, this turns $K$ into a discretely valued field.

\mi{When implementing valuations on a discrete valuation ring $R$ with field of fractions $K$, we made the design choice to put an instance of the valued class on $K$, but not on the ring $R$ itself. The reason is that, if needed, we can recover the
topological structure on $R$ from the one on $K$, so we prefer not to duplicate this information. This choice is consistent with \mathlib's file \url{ring_theory.dedekind_domain.adic_valuation}, in which valued instances are only put on fields.}

Both DVRs and valuations had already been defined in \fil{\mathlib SAY WHAT THIS IS SOMEWERE} before our project, but nothing beyong their basic properties was available. In particular, no definition of uniformisers or the relation between generators of $\maxid$ and the valuation on a DVR was available.

More crucially, two fundamental constructions were not formalised in \mathlib nor, \fil{to the best of our knowledge, in any other proof assistant CHECK}: extensions and completions of DVR. The next section is devoted to a description of our formalisation of these \fil{notions? constructions?}.

>>>>

%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[A Formalization of Complete DVRs and Local Fields]{A Formalization of Complete Discrete Valuation Rings and Local Fields}

@@

\section{Introduction}\label{sec:intro}
The vague idea that geometric intuition and algebraic rigor can fruitfully interact is an old theme, certainly predating the modern approach by Descartes and subsequently by Newton and Leibniz. 
In contemporary commutative algebra, and especially
after the advent of scheme theory by Grothendieck around 1960, this connection has become even tighter: geometric concepts and techniques are often borrowed for a wide range of applications. In this work we are concerned with the concept of \emph{local field}, a fundamental notion in algebraic number theory whose origin, and still many applications, comes from geometry. 
%If one is interested in studying arithmetic properties of the integers $\ZZ=\{\dots, -1,0,1,2,\dots\}$, the fact that every $n\in\ZZ$ can uniquely be factored as a product of prime numbers (up to reordering) is vital, as is the possibility of embedding $\ZZ$ inside the real numbers $\RR$, thus inheriting all metric properties of $\RR$.
Suppose, for instance, that one is interested in the integral solutions $(a_1,a_2,a_3)$ to $X_1^2+X_2^2+X_3^3=0$: clearly, since for positivity reasons the only \emph{real} solution is $(0,0,0)$, there cannot be any other integral solutions. On the other hand, $X_1^2+ X_2^2-3X_3^2=0$ certainly has real solutions, yet has no non-trivial solutions in $\FF[3]$. Now, an integral solution $(a_1,a_2,a_3)$ with $a_1\equiv a_2\equiv 0\pmod{3}$ must be $(0,0,0)$, as can be seen by comparing the occurrences of $3$ in the factorizations of $\alpha_1^2+\alpha_2^2$ and of $3\alpha_3^2$.

From a geometric perspective, one can regard the previous as a kind of ``local analysis'': interpreting the primes as the points of a geometric object attached to $\ZZ$, if a ``global'' solution $(a_1,a_2,a_3)\in\ZZ^3$ exists, we could obtain infinitely many solutions $(\overline{a_1},\overline{a_2},\overline{a_3})\in\FF^3$ ``around each point'', for varying $p$, as well as a solution $((a_1 : \RR),(a_2 : \RR),(a_3 : \RR))\in\RR^3$ around the ``point at infinity''. If at some point no solution can be found, then no global solution exists. Yet, in this analogy, the fields $\FF$ and $\RR$ are very different: the former is finite, with trivial discrete geometry and of positive characteristic, while the latter has a rich metric structure and characteristic $0$. Now, for every $p$, one can rather consider the field of $p$-adic numbers $\QQ_p$ which is, in this perspective, a better analogue for $\RR$ than the several $\FF$: it has a metric with respect to which it is complete (Cauchy sequences converge), and it has characteristic $0$ (so, in particular, it contains $\QQ$). To define it, observe that for every prime $p$ there exists an absolute value $\lvert\cdot\rvert_p$ on $\QQ$
% fix a prime number $p$: given any nonzero integer $n\in\ZZ$, we can factor it uniquely as $n=p^a\cdot n_0$, with $p\nmid n_0$. We then set $\lvert n\rvert_p = p^{-a}$, and $\lvert 0\rvert _p=0$. We can extend this function to $\QQ$ by declaring
% \[
% \left\lvert\frac{r}{s}\right\rvert_p = \frac{\lvert r\rvert_p}{\lvert s\rvert_p}.
% \]
% This function is actually an absolute value, 
for which numbers that are highly divisible by $p$ are closed to $0$. Having an absolute value, it is possible to speak about convergence and about Cauchy sequences: exactly as for the Euclidean absolute value, one can find Cauchy sequences that do not converge to any rational, and $\QQ_p$ is defined as the \emph{completion} of $\QQ$ with respect to $\lvert\cdot\rvert_p$. It is the ``smallest'' field where all $\lvert\cdot\rvert_p$-Cauchy sequences converge. Crucially, it still bears strong connections with the prime $p$: for instance, if a monic polynomial in $\ZZ[X]$ has a simple root in $\FF$, then it also has one in $\QQ_p$.

The local analysis we described before is an example of the ``local-to-global'' principle: to solve a problem in $\ZZ$, or in $\QQ$, one can try to solve it in $\QQ_p$, for all $p$, and in $\RR$. For a notorious example, it has long been known 
%that the Fermat \emph{equation} does not satisfy the local-to-global principle: \fil{add ex?}; yet, 
the so-called ``first case'' of Fermat's Last Theorem can be solved using class field theory (see~\cite{LenSte97}), which is a deep theory where the global theorems are obtained by a beautiful patching of the local ones: we refer to \cite{Tat67}~for an excellent account of this example. It should also be emphasized that ultimately Wiles' very proof of Fermat Last Theorem makes heavy use of the local-to-global principle in the 
%not by directly applying it to the equation itself (as we have seen that this brings no information), but in 
% the more general 
framework of Galois representations: we refer to~\cite{Fre09} for a summary of this technique.

Before describing our work, let us mention two generalizations of the $p$-adic numbers that will guide our approach to the formalization. First of all, the field $\QQ$ of rational numbers can be replaced by an arbitrary number field, or global field. We refer to~\cite[\S2]{BaaDahNarNuc22} for a brief overview of these notions in the context of formalization. Number fields are fields that can be obtained by adjoining roots of polynomials $f(X)\in\ZZ[X]$ to $\QQ$, like $\QQ(i)$ or $\QQ(\sqrt{2})=\{r+s\sqrt{2}, r,s\in\QQ\}$. They are a key object of study in algebraic number theory and resemble in many perspectives the rational field. Every number field $K$ can be realized as the field of fractions $K=\Frac{\rint{K}}$ of a suitable subring $\rint{K}\subseteq K$, in the same vein as $\QQ=\Frac{\ZZ}$. These rings $\rint{K}$ are not principal, in general, but they are Dedekind domains (see Example~\ref{example:adic_valuation} and the references~\ibid): so, although we cannot in general define an absolute value $\lvert\cdot\rvert_\pi$ on $\rint{K}$ associated to elements $\pi\in\rint{K}$, it is possible to define an absolute value $\lvert\cdot\rvert_{\maxid[\empty]}$ associated to every maximal ideal of $\rint{K}$. The completion procedure discussed above can then be performed analogously, and the fields $\compl[{\maxid[\empty]}]{K}$ obtained in this way are all finite extensions of some $\QQ_p$, and are the so-called mixed characteristic local fields. Analogous constructions exist when replacing $\QQ$ by $\FF(X)$ and we refer to~\S\ref{subsec:eq_char} for more details.

Secondly, unlike the Euclidean metric, the $p$-adic one carries a strong discrete flavor, in that only integral powers of $p$ can occur as values $\lvert x\rvert_p$ for $x\in\QQ$. This happens in other contexts and the relevant notion is that of a discrete valuation ring: we will describe their theory in more detail in~\S\ref{subsection:def_dvr}. Here we content ourselves with saying that they are a class of rings~$R$ endowed with a valuation map $v$ to $\ZZ$ such that the arithmetic behavior of elements $r\in R$ can be read on the value $v(r)$. The relation of discrete valuation rings with local fields is deep: the completion of the fraction field of a discrete valuation ring with finite residue field (see~\ref{sec:local_fields}) is a local field and, conversely, for every local field $F$ the subset $\{x \in F \vert \lvert x\rvert_F\leq 1\}$ is always a discrete valuation ring, yet not every discrete valuation ring is of this form.

In this paper, we formalize in \lean the definition and basic theory of local fields and their relations with discrete valuation rings. We provide both a concrete approach, defining mixed characteristic and equal characteristic local fields as finite field extensions of $\QQ_p$ and $\laurentseries$ and a more abstract definition of local field which comprises these two special cases.

The \mathlib library provides 
a very complete theory of general valuations, as well as the basic theory of discrete valuation rings, but the connection between these two notions is largely missing. We formalize the definition of discrete valuation and greatly expand the available theory for discrete valuation rings, their fields of fractions, and their associated discrete valuations, including theorems about completions, extensions of valuations and global-to-local results.
To the authors' best knowledge, this work constitutes the first formalization of local fields and of extensions of discrete valuations in any proof assistant.

The \lean[\empty] code for our formalization is available at a public anonymized GitHub repository\href{https://github.com/LCFT-Lean/local_fields}{\extlink}. The \code{from_mathlib} folder\href{https://github.com/LCFT-Lean/local_fields/tree/master/src/from_mathlib}{\extlink} contains files that are due to Mar\'ia In\'es de~Frutos-Fernández, Yaël Dillies and Filippo A.~E.~Nuccio as indicated in each file header and are used in this project after securing permission from them. Everything else is new, original work of the authors of this paper, totaling about 8,000 lines of code. Some of the \LClistingname s below have been edited for clarity, but we also provide links to the corresponding results in our repository. In some \LClistingname s the so-called \emph{dot notation} is exhibited: for example, the call \code{(ideal.is_prime I)} can be shortened to \code{I.is_prime}.

Throughout this paper, all rings are assumed to be commutative and unitary.

%%%%%%%%%%%%Subsec : Lean and mathlib
\subsection{\texorpdfstring{\lean[\empty] and \mathlib}{Lean and mathlib}}
This project is formalized in the \lean theorem prover \cite{Lean3}, which is based on dependent type theory, with proof irrelevance, quotient types, and non-cumulative universes \cite{TypeTheory}.

We build our work on top of the mathematical library \mathlib of \lean{\empty}, whose key properties are its unified approach to integrate different mathematical theories, and its decentralized nature, with over 300 contributors. One of the key tools for organizing mathematical hierarchies used both in \lean[\empty]'s core library and in \mathlib is typeclass inference. By declaring certain terms as \code{instance}, \lean[\empty] will automatically try to infer the relevant values during the unification procedure. We refer to~\cite{Baa22} for a more detailed discussion of the role of \code{instance} parameters in \lean[\empty].

The formalization is in \lean because at the time when we begun it almost none of it prerequisites were available in \lean[4]. Now that the complete \mathlib has been translated to \lean[4], we plan to port our work and start to integrate it in this library. All our future work related to this project will be directly implemented in \lean[4].

%%%%%%%%%%%%Subsection : paper outline
\subsection{Paper outline}
In \S\ref{section:dvr}, we treat the general theory of discrete valuations and discrete valuation rings: after presenting some background material in \S\ref{subsection:wzmZZ}--\S\ref{subsection:def_dvr}, we describe our main results in \S\ref{subsec:complete_DVR}, concerning completions, and \S\ref{subsec:extensions}, concerning extensions of valuations. In \S\ref{def:local_field} we present our formalization of local fields, describing the equal and mixed characteristic cases in \S\ref{subsec:eq_char} and \S\ref{subsec:mixed_char}, respectively. We conclude in \S\ref{sec:conclusion} with a description of future and related work and a discussion of some of our key design choices.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%Section : DVR
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section {Discrete Valuation Rings}\label{section:dvr}

In this section we present our formalization of the theory of discretely valued fields and discrete valuation rings, including results about completions and extensions of valuations. As a technical prerequisite, we discuss in \S~\ref{subsection:wzmZZ} the definition and main properties of the \mathlib type $\wzmZZ$, where most of the valuations that we considered take their values.

\subsection{\texorpdfstring{The type $\wzmZZ$}{The type with\_zero multiplicative Z}}\label{subsection:wzmZZ}
There are several situations in mathematics in which additive structures get translated into multiplicative structures. For example, associated to each real number $n \ne 0$, there is an exponential map $\exp_n : \RR \to \RR$ sending $x$ to $n^x$. This map has the property that
$\exp_n(x + y) = \exp_n(x) \cdot \exp_n(y)$ and, whenever $n > 1$, it preserves the order on $\RR$.

A second example occurs in the theory of valuations. If $\av : R \to \ZZ \cup \{\infty\}$ is an additive valuation on a ring $R$ (see Definition~\ref{def:add_valuation}) and $n > 1$ is a real number, it is common to study the associated function $v \colon R \rightarrow n^{\ZZ} \cup \{0\}$ sending $r$ to $n^{-\av(r)}$, with the convention that $n^{-\infty} = 0$. Here
\[ n^{\ZZ} \cup \{0\}=\{n^a \:\vert\: a \in \ZZ\} \cup \{0\}\subseteq \RR. \]

Some abstractions that can be used to formalize this kind of translations are available in \mathlib. Given any type \code{A} endowed with an additive structure, \mathlib provides a new type \code{multiplicative A} that is in bijection with \code{A} and carries a multiplicative structure, together with a map
\code{of_add} : \code {A} $\rightarrow$ \code{multiplicative A} satisfying
\[
    \mcode{of\_add\; (x + y) = of\_add\; x\; *\; of\_add\; y}
\]
for all \code{x, y : A}. If \code{A} is equipped with a preorder, then so is \code{multiplicative A}, and the map \code{of_add} is strictly monotone: \code{x} $\le$ \code{ y} is equivalent to \code{of_add x} $\le$ \code{of_add y}. The map \code{to_add : multiplicative A} $\to$ \code{A} is inverse to \code{of_add}.

We are especially interested in the case \code{A} $= \ZZ$. As an additive group, $\ZZ$ is an infinite cyclic group with generators~$1$ and~$-1$. Correspondingly, the elements \code{of_add(1)} and \code{of_add(-1)}, that do not bear a specific notation, are the only generators of the cyclic group \code{multiplicative}\;$\ZZ$. Since the map \code{of_add} preserves the unit element, it holds that \code{of_add(0) = 1} is the unit, hence
\[
\text{\code{(1:}}\;\text{\code{multiplicative} }\ZZ)\neq\text{\code{of_add(1:}}\ZZ\text{\code{)}}.
\]
Given two integers $a, b$, it holds that \code{of_add(a * b) = of_add(a)}$^b$\code{= of_add(b)}$^a$. In particular, we have the equality \code{of_add(}$\pm$\code{1)}$^n$= \code{of_add(}$\pm$\code{n)} for all \code{n :} $\ZZ$.

A new type \code{with_zero (multiplicative} $\ZZ$\code{)}, denoted $\wzmZZ$, can be constructed from \code{multiplicative} $\ZZ$ by adding an extra term \code{0} to \code{multiplicative} $\ZZ$. The resulting type $\wzmZZ$ still carries a multiplication, with extends the one on \code{multiplicative} $\ZZ$ and for which \code{0 * x = 0} for all \code{x}; moreover, setting \code{0} $\le$ \code{x} for all \code{x}, $\wzmZZ$ is endowed with an order that extends the one on \code{multiplicative} $\ZZ$. The inclusion
\[
\text{\code{multiplicative} }\ZZ \rightarrow \wzmZZ
\]
is an order-preserving coercion that respects the multiplication on both sides, and we omit it from the notation.

The type $\wzmZZ$ provides an abstraction of the structure on the set $n^{\ZZ} \cup \{0\}$ that does not require a choice of the base $n$, and it will be the codomain of most valuations that we consider in this paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%Subsection : Valuations
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Valuations}\label{subsection:valuations}
\begin{definition}[{\cite[Chapitre~VI, \S3, n$^\circ$1, D\'efinition~1]{Bou85}}]\label{def:add_valuation}
Let $R$ be a ring. An \emph{additive valuation} on $R$ is a function $\av \colon R \to \ZZ \cup \{\infty\}$ such that
\begin{listDef}
	\item $\av(r) = \infty$ if and only if $r = 0$; \label{def_add_valuation:zero}
	\item $\av (r\cdot s) = \av(r) + \av(s)$ for all $r, s$ in $R$;\label{def_add_valuation:mul}
	\item $\av (r + s) \geq \min \{\av(r), \av(s)\}$ for all $r, s$ in $R$. \label{def_add_valuation:add}
\end{listDef}
\end{definition}

Given an additive valuation $\av$ on $R$, we can define an associated function $v \colon R \rightarrow \wzmZZ$ by sending $0$ to $0$ and $r \ne 0$ to \code{of_add(-}$\av$\code{(r))}. From the definition of $\av$, it is easy to deduce that $v$
satisfies the properties of a multiplicative valuation, as in the following definition:
\begin{definition}\label{def:valuation}
A \emph{multiplicative valuation} on a ring $R$ is a function $v\colon R \rightarrow \wzmZZ$ satisfying the properties
\begin{listDef}
	\item $v (0) = 0$;\label{def_valuation:zero}
	\item $v (1) = 1$;\label{def_valuation:one}
	\item $v (x \cdot y) = v (x) \cdot v (y)$ for all $x, y \in R$;
 \label{def_valuation:mul}
	\item $v (x + y) \le \max \{v (x), v (y)\}$ for all $x, y \in R$.\label{def_valuation:add}
 \end{listDef}

\end{definition}
Note that an element $r \in R$ has additive valuation $1$ if and only if it has multiplicative valuation \code{of_add(-1)}. Elements of multiplicative valuation \code{of_add(-1)} will play a prominent role in \S\ref{subsection:def_dvr}.

For example, let $R = \ZZ$ be the ring of integers, and fix a prime number $p$. Then, thanks to unique factorization, we can define an additive valuation $\av[p] \colon \ZZ \to \ZZ \cup \{\infty\}$ by sending an integer $z$ to the number of times that $p$ appears in the factorization of $z$. We can extend the function $\av[p]$ to $\QQ$ by defining $\av[p](r/s) := \av[p](r) - \av[p](s)$. Then $\av[p]$ is an additive valuation on $\QQ$, called the \emph{additive $p$-adic valuation} on $\QQ$. 
The corresponding multiplicative valuation $v_p \colon \QQ \to \wzmZZ$ is called the \emph{$p$-adic valuation}.

% The linear order on $\wzmZZ$ allows us to compare elements of a ring endowed with a valuation : for example, the integers $20$ and $25$ have $5$-adic valuation equal to \code{of_add (-1)} and \code{of_add(-2)} in  $\wzmZZ$, respectively, so $25$ is ``smaller'' than $20$ with respect to the $5$-adic valuation.

\begin{example}[The {$\maxid[\empty]$}-adic valuation]\label{example:adic_valuation}
We will often consider the following generalization of the $p$-adic valuation. If $R$ is a Dedekind domain (see~\cite[Chapitre~VII, \S2]{Bou85}) which is not a field, then every nonzero ideal of $R$ can be factored as a product of maximal ideals, uniquely up to reordering. Therefore, for every maximal ideal $\maxid[\empty]$ of $R$, we can follow an analogous construction to define an additive valuation $\av[{\maxid[\empty]}] \colon R \to \ZZ \cup \{ \infty\}$ on $R$ by sending $0$ to $\infty$ and any nonzero $r \in R$ to the number of times that $\m$ appears in the factorization of the principal ideal $(r)$. The corresponding multiplicative valuation $v_{\maxid[\empty]}=\mcode{of\_add}\circ (-\av[{\maxid[\empty]}])\colon K \to \wzmZZ$ on the fraction field $K$ of $R$ is called the \emph{$\m$-adic valuation}. This valuation was formalized in \cite{deF22}, and is available in \mathlib as the declaration~\code{is_dedekind_domain.height_one_spectrum.valuation}\href{https://leanprover-community.github.io/mathlib_docs/ring_theory/dedekind_domain/adic_valuation.html#is_dedekind_domain.height_one_spectrum.valuation}{\extlink}.
\end{example}

While in most number theory references it is more common to work with additive valuations than with multiplicative ones, for historical reasons \mathlib follows the opposite convention, and it provides a much more complete API for multiplicative valuations. Hence we use multiplicative valuations in our formalization and, throughout the paper, whenever we use the term "valuation" without further adjectives, we mean "multiplicative valuation".

\begin{remark}\label{rmk:value_group}(For experts)
More generally, the codomain $\wzmZZ$ in the definition of a (multiplicative) valuation can be replaced by $\Gamma_0=\Gamma\cup\{0\}$, where $\Gamma$ is a linearly ordered commutative group. The order on $\Gamma$ is extended to $\Gamma_0$ analogously as we did for $\wzmZZ$.
